En Garde! Unguarded Iteration for Reversible Computation in the Delay Monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-06-28 mpc19/paper.tex
Some corrections as suggested by the reviewers.

\Pio{} is a family of reversible combinator calculi comprising structural
isomorphisms and combinators corresponding to those found in dagger-traced
$\omega$-continuous rig categories~\cite{Karvonen19}. Though superficially
simple, \Pio{} is expressive enough as a metalanguage to give semantics to the
typed reversible functional programming language Theseus~\cite{JamesS14}.

In \cite{CaretteS16}, the authors studied the equivalences of isomorphisms in
the finitary fragment of \Pio{} (i.e., without recursive types and iteration
via the trace combinator), and showed that these equivalences could be
adequately described by another combinator calculus of equivalences of
isomorphisms, in sum yielding a two-level calculus of isomorphisms and
equivalences of isomorphisms. In this paper, we build on this work to produce a
(fully formalized) two-level calculus for full \Pio{} (supporting both recursive
types and iteration) via the delay monad, using insights gained from the
study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as well
as of join inverse categories in which reversible iteration may be
modelled~\cite{KaarsgaardAG17}.

@@

It is possible to show that every term $f$ has $\dagger\,f$ as its partial
inverse, and since the final equivalence of Figure~\ref{fig:programequivs} is
equivalent to stating that partial inverses are unique~\cite{Kastl79}, it
follows that the types and terms of \Pio\ form an inverse category. The proof
proceeds by induction on $f$.
\[
\mathsf{existsPIso} : (f : A \lr B) \to f \bullet \dagger\,f
\bullet\,f \LR f
\]

@@

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is
not the
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. 
%The binary product of $A$ and $B$ in $\Dapprox$ is given by $A + B + A \times B$. 
Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

@@

Summing up, in this section we have showed that the $\traceD$ operator
can be restricted to act on partial isomorphisms. That is, the
following type is inhabited:
\[
\semTrace : A + C\simeq B + C \to A \simeq C
\]

>>>>

\Pio{} is a family of reversible combinator calculi comprising structural
isomorphisms and combinators corresponding to those found in dagger-traced
$\omega$-continuous rig categories~\cite{Karvonen19} (a kind of dagger category
with a trace, monoidal sums $\oplus$ and products $\oplus$ such that they form
a rig structure, and fixed points of the functors formed from the rig
structure). Though superficially simple, \Pio{} is expressive enough as a
metalanguage to give semantics to the typed reversible functional programming
language Theseus~\cite{JamesS14}.

In \cite{CaretteS16}, Carette and Sabry studied the equivalences of 
isomorphisms in the finitary fragment of \Pio{} (i.e., without recursive types
and iteration via the trace combinator), and showed that these equivalences
could be adequately described by another combinator calculus of equivalences of
isomorphisms, in sum yielding a two-level calculus of isomorphisms and
equivalences of isomorphisms. In this paper, we build on this work to produce a
(fully formalized) two-level calculus for full \Pio{} (supporting both
recursive types and iteration) via the delay monad, using insights gained from
the study of its Kleisli category~\cite{UustaluV17,ChapmanUV19,VeltriPhD}, as
well as of join inverse categories in which reversible iteration may be
modelled~\cite{KaarsgaardAG17}.

@@

It is possible to show that every term $f$ has $\dagger\,f$ as its partial
inverse. The notion of partial inverse used here comes from the study of
\emph{inverse categories} (see \cite{Kastl79}) and amounts to saying that
$\dagger\,f$ is the \emph{unique} map that undoes everything which $f$ does
(unicity of partial inverses follows by the final equivalence of
Figure~\ref{fig:programequivs}, see~\cite{Kastl79}). Note that this is
different from requiring that $f$ is an isomorphism in the usual sense, as
$\dagger\,f \bullet f$ is not going to be the identity when $f$ is only
partially defined, though it will behave as the identity on all points where
$f$ is defined.

The proof that every term has $\dagger\,f$ as its partial inverse proof
proceeds by induction on $f$.
\[
\mathsf{existsPIso} : (f : A \lr B) \to f \bullet \dagger\,f
\bullet\,f \LR f
\]

@@

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories, a kind of
categories of partial maps) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is
not the
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. 
%The binary product of $A$ and $B$ in $\Dapprox$ is given by $A + B + A \times B$. 
Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

@@

Summing up, in this section we have showed that the $\traceD$ operator
can be restricted to act on partial isomorphisms. That is, the
following type is inhabited:
\[
\semTrace : A + C\simeq B + C \to A \simeq B
\]
